---
title: "Lab 4"
output: html_document
date: "2024-10-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import libraries

```{r}
library(here)
library(lubridate)
library(readxl)
library(janitor)
library(tidyverse)
library(sf)
```

```{r}
electricity_district <- read_csv("studentData/electricity_district.csv")
head(electricity_district)
```

```{r}
dictionary <- read_csv("studentData/dictionary.csv")
head(dictionary)
```

# Exploring the data

```{r}
dictionary |>
  filter (name == "Hinton / Edson")
```

```{r}
population <- read_csv("studentData/population.csv")
head(population)
```

# Plotting using shape files

```{r}
planning_areas <- st_read("studentData/shape/AESO_Planning_Areas.shp") |>
  mutate(Area_ID = as.numeric(Area_ID))
head(planning_areas)
```

Let's build a basic map to understand the capabilities of a shape file.

```{r}
# Basic plot
base_map <- ggplot(planning_areas) +
  geom_sf() +    
  labs(title = "Alberta") 
base_map
```

Perfect! Now we want to fill this map with data. For the moment, we only care about 9 districts given in the dataset "districts_included.csv".

```{r}
districts_included <- read.csv("studentData/districts_included.csv") 
districts_included
```

In order to plot those 9 district, the first approach is to conduct an inner join, assuming we don't care about the rest of the districts.

```{r}
planning_areas_trial <- planning_areas %>%
  inner_join(districts_included, by = "Area_ID")
head(planning_areas_trial)
```

Let's plot this new data frame (the output of our inner join).

```{r}
alberta <- ggplot(planning_areas_trial) +
  geom_sf(fill = "lightblue", color = "black") +    
  geom_sf_text(data = st_centroid(planning_areas_trial),             
               aes(label = Area_ID),                   
               color = "darkred", size = 3) +       
  labs(title = "Alberta") 
alberta
```

This map, while focusing on just our problem at hand, is not a complete solution as it does not show the entire map. This output is not desirable. Let's make some small changes in the data to obtain the full picture.

```{r}
planning_areas <- planning_areas %>%
  left_join(districts_included, by = "Area_ID")
head(planning_areas)
```

```{r}
alberta_complete <- ggplot(planning_areas) +
  geom_sf(fill = "lightblue", color = "black") +    
  geom_sf_text(data = st_centroid(planning_areas),             
               aes(label = Area_ID),                   
               color = "darkred", size = 3) +       
  labs(title = "Alberta") 
alberta_complete
```

Out of this, we only want to label the 9 districts :

|                          |             |
|--------------------------|-------------|
| **Name**                 | **Area_ID** |
| Lethbridge               | 54          |
| Athabasca / Lac La Biche | 27          |
| Vegreville               | 56          |
| High Level               | 18          |
| Alliance / Battle River  | 36          |
| Brooks                   | 47          |
| Grande Prairie           | 20          |
| Wabamun                  | 40          |
| Hinton / Edson           | 29          |

So, we will create a new column in "planning_areas" called id_label that returns the Area_ID if the Area_ID is included in the 9 districts and NA if it is not.

One idea to do this is to create the id_label column in "districts_included" and then left join it to planning areas.

```{r}
districts_included <- districts_included |>
  mutate(id_label = Area_ID) |>
  select(Area_ID, id_label)
districts_included
```

```{r}
planning_areas_joined <- planning_areas |>
  left_join(districts_included, by = "Area_ID") |>
  mutate(id_label = as.character(id_label)) |>
  mutate(id_label = ifelse(is.na(id_label), "", id_label))
head(planning_areas_joined)
```

Now, as you can see, id_label has empty rows for the districts we aren't concerned about. Let's plot this data frame with id_label as the labels.

```{r}
st_centroid(planning_areas_joined)
```

```{r}
alberta_9_dist <- ggplot(planning_areas_joined) +
  geom_sf(fill = "lightblue", color = "black") +    
  geom_sf_text(data = st_centroid(planning_areas_joined),             
               aes(label = id_label),                   
               color = "darkred", size = 3) +       
  labs(title = "Alberta ") 
alberta_9_dist
```

We have finally created a desirable chart. Let's attempt some visualizations.

Problem statement: Visualize the relative energy consumption in the 9 districts.

For this, we need the population of the 9 districts, their respective energy consumption and consumption per individual (compare apples to apples).

```{r}
electricity_district <- electricity_district |>
  rename("18" = AREA18,
         "20" = AREA20,
         "27" = AREA27,
         "29" = AREA29,
         "36" = AREA36,
         "40" = AREA40,
         "47" = AREA47,
         "54" = AREA54,
         "56" = AREA56)
head(electricity_district)
```

```{r}
electricity_long <- electricity_district %>%
  pivot_longer(cols = c("18","20","27","29","36","40","47","54","56"),         
               names_to = "Area_ID",       
               values_to = "electricity_use") 
head(electricity_long)
```

We want to get the average energy consumption for each Area_ID

```{r}
mean_electricity <- aggregate(electricity_use ~ Area_ID, 
                              data = electricity_long, FUN = mean)

mean_electricity <- mean_electricity |>
   mutate(Area_ID = as.numeric(Area_ID))

print(mean_electricity)
```

```{r}
# We're only working with 2011

population_2011 <- population |>
  filter(year == 2011) |>
  mutate(Area_ID = code) |>
  mutate(Area_ID = as.numeric(Area_ID))
population_2011
```

```{r}

elec_pop <- mean_electricity |>
  left_join(population_2011, by = "Area_ID")
elec_pop
```

Create a column for per capita consumption.

```{r}
elec_pop_final <- elec_pop |>
  mutate(per_capita = electricity_use/population) 

elec_pop_final
```

In order to plot this, we need to bring back the shape data, and pre-process such that only the 9 districts are plotted.

```{r}
shape_viz <- planning_areas_joined |>
  left_join(elec_pop_final, by = "Area_ID")
head(shape_viz)
```

```{r}
ggplot() +
  geom_sf(data = shape_viz,
          aes(fill = ifelse(per_capita > 0, per_capita, NA))) +
  # scale_fill_viridis_c(na.value = "gray90") +      
  labs(title = "Electricity consumption per capita",
       fill = "Consumption") 
```

To make this better, let's also add the labels (of each district) back to this plot.

```{r}
ggplot() +
  geom_sf(data = shape_viz,
          aes(fill = ifelse(per_capita > 0, per_capita, NA))) +
  scale_fill_viridis_c(na.value = "gray90") +  
  geom_sf_text(data = st_centroid(planning_areas_joined),             
               aes(label = id_label),                   
               color = "white", size = 3) +
  labs(title = "Electricity consumption per capita",
       fill = "Consumption") 
```
